Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-10-03 analysis/paper/paper.Rmd
Various - modifying methods and results...

We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships with
available local demographic data in a simple Bayesian population model
for boreal caribou. In our simple proof-of-concept model, information is 
shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020]. A more
rigorous statistical approach would be to integrate multiple populations
into a hierarchical Bayesian analysis of variation within and among
populations [@leasure_hierarchical_2019]; we have not done this because
owners have not authorized the use of the data for this purpose,
obtaining permission to use caribou demographic data is difficult, and
most managers only have access to their own local data. Despite
limitations, our simple Bayesian model represents an important
improvement over less formally integrated methods for estimating and
projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, improve viability projections on changing
landscapes, and help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach can help clarify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

@@

resultScenario = "S10"
resultSet ="cmult6ay3aSf1"
figSet = c("diffs","distScns","EVSI","power","EVPIall")

@@

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios. Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%). These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, and not intended to reflect real landscape change. In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))

@@

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow
ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification biases
(Supplement B) and (optionally) a delayed reproduction term to account
for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum
potential recruitment of female calves per cow $\dot{X}_t$:

\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}

@@

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$). Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

knitr::include_graphics(here::here("analysis/paper/figs/DemographicRatesDelay.png"))

@@

To account for bias and additional uncertainty caused by unknown
misidentification biases (Supplement B) and delayed age of age of first
reproduction [@eacker_web-based_2019; @decesare_estimating_2012] the
estimated recruitment probability is adjusted to get expected female
recruits per female: 
\begin{equation}
X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
  (\#eq:X)
\end{equation}

The composition survey bias correction term $c$ depends on the known
apparent number of adult females per collared animal $w$, the unknown
ratio of young bulls to adult females $q$, the unknown probability of
misidentifying young bulls as adult females and vice versa $u$, and the
unknown probability of missing a calf $z$ (Supplement B). Prior
uncertainty about the value of the bias correction term $c$ is
approximated with a Log-normal distribution (Supplement B). Given the
apparent number of adult females per collared animal $w$ in each
monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate
the mean and standard deviation of 10,000 samples of $\log{ c}$ assuming
the ratio of young bulls to adult females $q$ varies uniformly between 0
and 0.6, the adult misidentification probability $u$ varies uniformly
between 0 and 0.2, and the probability of missing a calf $z$ varies
uniformly between 0 and 0.2. To maintain consistency with Johnson et al.
[-@johnson_science_2020] we also provide an option to only adjust for
sex ratio and misidentification biases $X_t=cR_t/2$.

@@

As in the simulations, we assume a simple demographic model, with a
census in late winter of surviving females and new recruits to the
post-juvenile (aged 1+) class. The number of post-juvenile females that
survive from year $t$ to the census $W_t$ is binomially distributed with
survival probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The
number of juveniles recruiting to the post-juvenile class at the census
is a Poisson distributed function of the number of surviving
post-juvenile females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Poisson}(W_tX_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y=3$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 5000 animals. For reference, all
variables and parameters are defined in supplemental Table A1.

```{=html}
<!-- Annotated equations for poster.
$$J_{t} \sim \text{Poisson}(W_tX_t)$$
$$\underbrace{N_{t+1}}_{\text{Adult females in year t+1}}=\underbrace{W_t}_{\text{Surviving adults}}+\underbrace{J_t}_{\text{New recruits to adult age class}}$$
$$W_{t} \sim \text{Binomial}(N_t,\underbrace{S_t}_{\text{Adult survival probability}})$$
$$S_t=\underbrace{\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})}_{\text{Adjustment as in in Johnson et al. 2020}}=\beta^S_0+ \underbrace{\beta^S_aA_t}_{\text{Anthropogenic disturbance effect}}+\underbrace{\epsilon^S_t}_{\text{Random effect of year}}.$$
$$J_{t} \sim \text{Poisson}(W_t\underbrace{R_t/2}_{\text{Recruitment rate}})$$
$$\log(R_t)=\beta^R_0+\underbrace{\beta^R_a A_t}_{\text{Anthropogenic disturbance effect}}+ \underbrace{\beta^R_fF_t}_{\text{Fire effect}}+\underbrace{\epsilon^R_t}_{\text{Random effect of year}}.$$
$$\underbrace{\hat{C}_t}_{\text{Observed calves}} \sim \text{Binomial}(\underbrace{\hat{W}_t}_{\text{Observed cows}},R_t)$$
$$\underbrace{\hat{S_t}}_{\text{Observed survival (Kaplan-Meier estimate)}} \sim \text{Normal}(S_t,1/\tau_t).$$ -->
```
```{=html}
<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives biased results if there is left censoring, or months without animals at risk. To address this we drop left censored observations and only estimate survival when at least some animals remain at risk by month 12.
\*
see alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. See also Lamb et al, who uses two step KM estimation process for survival like Eacker. Stick with published examples. \*
-->
```
Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with high (b: $v_R=0.46,v_S=0.087$) and lower (c: $v_R=0.23,v_S=0.0435$) interannual variation. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty more when interannual variation is lower (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99). When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}

@@

```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations. In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
```

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

@@

knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay3aSf1.png"))
```

@@

The main objective of Johnson et al's [-@johnson_science_2020;
@eccc_scientific_2011] national analysis was to inform a national
disturbance policy that applies to all ranges except SK1, and to inform
a different policy for SK1 [@eccc_scientific_2011; @eccc_boreal_2019]. A
hierarchical analysis [@leasure_hierarchical_2019] would be better
suited for estimating variation in demographic model parameters among
populations, and distinguishing that from variation among years.
However, data owners have not authorized the use of the data for that
purpose. To develop a proof-of-concept model with available published information we
constrained our example trajectory simulations by assuming that
populations remain in their quantile, so populations that are lower than
or higher than average remain lower or higher than average over time.
The Bayesian model is not similarly constrained, and can accommodate
evidence of a shift from unexpectedly low to unexpectedly high
demographic rates over time, or vice versa. Switching from low to high
each year is unlikely, but there is evidence of variation in the slope
of demographic-disturbance relationships among populations
[@rudolph_demographic_2017] so it is important that the Bayesian model
allows for a greater range of possibilities than our more constrained
simulated example trajectories. Although our simulated trajectories do
not include all plausible possibilities we believe they are sufficient
to demonstrate how a Bayesian model can be used to integrate local data
with national priors, and to show how monitoring requirements depend on
landscape condition.

For this analysis we used Kaplan-Meier estimates of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ [as in @eacker_web-based_2019; @lamb_effectiveness_2024], and assumed Beta distributed interannual variability [as in @johnson_science_2020], but we acknowledge there are other possible approaches [e.g. @dalgarno_bboutools_2024]. We believe the proof-of-concept model presented in this paper is adequate for making general points about the integration of national
demographic-disturbance relationships with local data, and accounting for
composition survey errors. However, in future we intend to use the more fully developed and documented 'bboutools' R package [@dalgarno_bboutools_2024] as a starting point for work with local decision makers to develop tools they can use inform local monitoring decisions.

Aspatial demographic models generally assume that observed animals are a
random and representative sample of a fully mixed population in which
survival and recruitment rates apply equally to all animals. However,
metrics derived from fecal pellet analysis such as reproductive success
[@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;
@mcfarlane_identification_2022], age structure and pregnancy rates
[@flasko_fecal_2017; @mcfarlane_identification_2022], and
mother-offspring co-occurence [@jones_novel_2023] suggest high variation
in fitness among individuals [@mcfarlane_genetic_2018] and spatial
structure within populations [@jones_novel_2023;
@mcfarlane_spatial_2021; @mcfarlane_identification_2022]. Analyses of
movement data also show spatial variation in risks and benefits for
caribou and variation in habitat preferences and space use among
individuals [e.g. @decesare_linking_2014; @avgar_spaceuse_2015].
Although there is evidence of spatial structure and variation in
demographic rates among individuals within populations, for most caribou
populations there is not enough data available to estimate spatial
variation in demographic model parameters.

A number of additional complexities can be incorporated into aspatial
Bayesian population models, including variation in demographic rates
among individuals [@armstrong_incorporating_2021], population size
information from aerial or fecal dna surveys, and additional age classes
[@moeller_integrating_2021; @mcnay_demographic_2022]. However, in many
contexts, population size and fecal dna survey information is not
available, and a framework that includes these sources of information
will be unnecessarily complicated. At present, our model is intended as
a proof-of-concept to show the value of integrating national
demographic-disturbance relationships with local data, accounting for
composition survey errors, and assessing the value of local monitoring
information. Rather than further complicate a proof-of-concept model
applied to simulated data, our next step will be to collaborate with an
interested decision maker to develop a tool informed by this analysis
that meets their real local needs, recognizing that these needs will
vary among monitoring programs [@nbckc_nbckc_2021; @nbckc_nbckc_2021].

## Assessing the value of additional monitoring

We applied the framework proposed by Dunham et al. (2023) to show that
no additional local monitoring information is needed to predict whether
boreal caribou populations in highly disturbed landscapes are
self-sustaining. There is more variation in caribou demographic rates
among study areas with low to moderate disturbance, and thus more need
for local data to inform demographic projections. In these places, our
analysis suggests that there may be diminishing returns from
increasing monitoring effort, but more information on monitoring costs,
objectives, and existing local data would be required to assess the
adequacy and cost-effectiveness of monitoring scenarios in particular
circumstances. Monitoring costs vary substantially with technologies
used, the remoteness of herds, and the difficulty of collaring animals,
which in turn depends on the number of currently collared animals, the
population size, and vegetation across the landscape
[@nbckc_boreal_2021]. It is important to recognize that projecting
population viability is only one possible objective for monitoring
[@nbckc_boreal_2019], and that the impacts of monitoring on threatened
animals should be considered [@nbckc_boreal_2019]. Since monitoring
costs, options and goals are context specific, we believe the most
effective way forward will be to use this proof-of-concept model and
analysis as a starting point for working with local decision makers to
build decision support tools suited to their particular circumstances,
needs and questions.

>>>>

We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships with
available local demographic data in a simple Bayesian population model
for boreal caribou. In our simple proof-of-concept model, information is 
shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020]. The model represents an important improvement over less formally integrated methods for estimating and
projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, improve viability projections on changing
landscapes, and help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach can help clarify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

@@

resultScenario = "S10"
resultSet ="cmult6ay5aSf1"
figSet = c("diffs","distScns","EVSI","power","EVPIall")

@@

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios. Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%). These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, and not intended to reflect real landscape change. In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay5aSf1.png"))

@@

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow
ratio $\dot{R}_t$ is adjusted for sex ratio and misidentification biases
(Supplement B) to get maximum
potential recruitment of female calves per cow $\dot{X}_t$:

\begin{equation}
  \dot{X}_t=\dot{c}\dot{R}_t/2;\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}

@@

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships. To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$). Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models.", cache=TRUE}

knitr::include_graphics(here::here("analysis/paper/figs/DemographicRates.png"))

@@

To account for sex ratio of calves and additional uncertainty caused by unknown
misidentification errors (Supplement B) the
estimated recruitment probability is adjusted to get expected female
recruits per female: 
\begin{equation}
X_t=cR_t/2; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
  (\#eq:X)
\end{equation}

The composition survey bias correction term $c$ depends on the known
apparent number of adult females per collared animal $w$, the unknown
ratio of young bulls to adult females $q$, the unknown probability of
misidentifying young bulls as adult females and vice versa $u$, and the
unknown probability of missing a calf $z$ (Supplement B). Prior
uncertainty about the value of the bias correction term $c$ is
approximated with a Log-normal distribution (Supplement B). Given the
apparent number of adult females per collared animal $w$ in each
monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate
the mean and standard deviation of 10,000 samples of $\log{ c}$ assuming
the ratio of young bulls to adult females $q$ varies uniformly between 0
and 0.6, the adult misidentification probability $u$ varies uniformly
between 0 and 0.2, and the probability of missing a calf $z$ varies
uniformly between 0 and 0.2.

@@

As in the simulations, we assume a simple demographic model, with a
census in late winter of surviving females and new recruits to the
post-juvenile (aged 1+) class. The number of post-juvenile females that
survive from year $t$ to the census $W_t$ is binomially distributed with
survival probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The
number of juveniles recruiting to the post-juvenile class at the census
is a binomially distributed function of the number of surviving
post-juvenile females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Binomial}(W_t,X_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y=3$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 5000 animals. For reference, all
variables and parameters are defined in supplemental Table A1.

```{=html}
<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives biased results if there is left censoring, or months without animals at risk. To address this we drop left censored observations and only estimate survival when at least some animals remain at risk by month 12.
\*
see alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. See also Lamb et al, who uses two step KM estimation process for survival like Eacker. Stick with published examples. \*
-->
```
Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty more when interannual variation is lower (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99). When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}

@@

```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations. In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay5aSf1.png"))
```

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay5aSf1.png"))
```

@@

knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay5aSf1.png"))
```

@@

The main objective of Johnson et al's [-@johnson_science_2020;
@eccc_scientific_2011] national analysis was to inform a national
disturbance policy that applies to all ranges except SK1, and to inform
a different policy for SK1 [@eccc_scientific_2011; @eccc_boreal_2019]. A
hierarchical analysis [@leasure_hierarchical_2019] would be better
suited for estimating variation in demographic model parameters among
populations, and distinguishing that from variation among years.
We have not done this because data owners have not authorized the use of the data for that purpose, and obtaining permission to use caribou demographic data is challenging. To move ahead with available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than or higher than average remain lower or higher than average over time.
The Bayesian model is not similarly constrained, and can accommodate
evidence of a shift from unexpectedly low to unexpectedly high
demographic rates over time, or vice versa. Switching from low to high
each year is unlikely, but there is evidence of variation in the slope
of demographic-disturbance relationships among populations
[@rudolph_demographic_2017] so it is important that the Bayesian model
allows for a greater range of possibilities than our more constrained
simulated example trajectories. Although the simulated trajectories do
not include all plausible possibilities we believe they are sufficient
to demonstrate how a Bayesian model can be used to integrate local data
with national priors, and to show that monitoring requirements depend on
landscape condition.

For this analysis we used Kaplan-Meier estimates of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ [as in @eacker_web-based_2019; @lamb_effectiveness_2024], and assumed Beta distributed interannual variability [as in @johnson_science_2020], but we acknowledge there are other possible approaches [e.g. @dalgarno_bboutools_2024]. The proof-of-concept model presented in this paper is adequate for making general points about the integration of national
demographic-disturbance relationships with local data, and accounting for
composition survey errors. However, the implementation is a research prototype, not a fully developed analysis tool. The more fully developed and documented *bboutools* R package [@dalgarno_bboutools_2024] is a better starting point for development of tools to enable local decision makers to inform decisions with their  existing data. A possible way forward would be to extend that package to include demographic-disturbance relationships and composition survey errors.

Aspatial demographic models generally assume that observed animals are a
random and representative sample of a fully mixed population in which
survival and recruitment rates apply equally to all animals. However,
metrics derived from fecal pellet analysis such as reproductive success
[@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;
@mcfarlane_identification_2022], age structure and pregnancy rates
[@flasko_fecal_2017; @mcfarlane_identification_2022], and
mother-offspring co-occurence [@jones_novel_2023] suggest high variation
in fitness among individuals [@mcfarlane_genetic_2018] and spatial
structure within populations [@jones_novel_2023;
@mcfarlane_spatial_2021; @mcfarlane_identification_2022]. Analyses of
movement data also show spatial variation in risks and benefits for
caribou and variation in habitat preferences and space use among
individuals [e.g. @decesare_linking_2014; @avgar_spaceuse_2015]. It is possible to integrate complexities such as variation in demographic rates
among individuals [@armstrong_incorporating_2021], population size
information from aerial or fecal dna surveys, and variation in demographic rates among age classes [@moeller_integrating_2021; @mcnay_demographic_2022] into Bayesian population models. However, in many contexts there is not enough to data to estimate more complex model parameters.

At present, our analysis is intended as a proof-of-concept to show the value of integrating national demographic-disturbance relationships with local data, accounting for
composition survey errors, and assessing the value of local monitoring
information. We applied the framework proposed by Dunham et al. (2023) to show that
no additional local monitoring information is needed to predict whether
boreal caribou populations in highly disturbed landscapes are
self-sustaining. There is more variation in caribou demographic rates
among study areas with low to moderate disturbance, and thus more need
for local data to inform demographic projections. In these places, our
analysis suggests that there may be diminishing returns from
increasing monitoring effort, but more information on monitoring costs,
objectives, and existing local data would be required to assess the
adequacy and cost-effectiveness of monitoring scenarios in particular
circumstances. Monitoring costs vary substantially with technologies
used, the remoteness of herds, and the difficulty of collaring animals,
which in turn depends on the number of currently collared animals, the
population size, and vegetation across the landscape
[@nbckc_boreal_2021]. It is important to recognize that projecting
population viability is only one possible objective for monitoring
[@nbckc_boreal_2019], and that the impacts of monitoring on threatened
animals should be considered [@nbckc_boreal_2019]. Since monitoring
costs, options and goals are context specific, we believe the most
effective way forward will be to use this proof-of-concept analysis, existing local data, and the more fully developed *bboutools* R package [@dalgarno_bboutools_2024] as a starting point for working with interested decision makers to develop monitoring decision support tools that meet real local needs, recognizing that those needs will vary [@nbckc_nbckc_2021;@nbckc_nbckc_2022].

### 2024-10-03 analysis/paper/supplement.Rmd
Various - modifying methods and results...

|$X_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|

@@

|$\dot{X}_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|

@@

If we also adjust for delayed age at first reproduction (DeCesare et al. 2012; Eacker et al. 2019), the adjusted recruitment rate becomes $$X=\frac{cR/2}{1+cR/2}.$$

>>>>

|$X_t$|rate|Recruitment adjusted for sex ratio and misidentification bias.|

@@

|$\dot{X}_t$|rate|Recruitment adjusted for sex ratio and misidentification bias.|

@@
